---
title: "Visualization II with ggplot2"
author: "Maliha Safdar"
date: "2025-09-30"
output: github_document
---

```{r}
library(tidyverse)
library(patchwork)
```
`We’ll still work with NOAA weather data, which is loaded using the same code as in Visualization Pt 1.`


```{r}
library(p8105.datasets)
data(weather_df)
```


`Revist the scatterplot tmax against tmin made in Visualization I.`

```{r}
weather_df |> 
  ggplot(aes(x = tmin, y = tmax)) + 
  geom_point(aes(color = name), alpha = .5)
```


### Labels

`One of the most important things you can do is provide informative axis labels, plot titles, and captions, all of which can be controlled using labs().`

```{r}
weather_df |> 
  ggplot(aes(x = tmin, y = tmax)) + 
  geom_point(aes(color = name), alpha = .5) + 
  labs(
    title = "Temperature plot",
    x = "Minimum daily temperature (C)",
    y = "Maxiumum daily temperature (C)",
    color = "Location",
    caption = "Data from the rnoaa package",
    color = "Location"
  )
```


### Scales

`you’ll occasionally want control over the location and specification of tick marks on the X or Y axis. These can be manipulated in scale_x_* and scale_y_* where * depends on the type of variable mapped to the x and y aesthetics (i.e. continuous vs discrete).`

```{r}
weather_df |> 
  ggplot(aes(x = tmin, y = tmax)) + 
  geom_point(aes(color = name), alpha = .5) + 
  labs(
    title = "Temperature plot",
    x = "Minimum daily temperature (C)",
    y = "Maxiumum daily temperature (C)",
    color = "Location",
    caption = "Data from the rnoaa package") + 
  scale_x_continuous(
    breaks = c(-20, 0, 25), 
    labels = c("-20º C", "0", "25")) +
  scale_y_continuous(
    trans = "sqrt",
    limits = c(10,30)
  )
```

`for scale_x_ continuos, continuos is chosen because we are changing the x-axis.`


`Using color hues.`

```{r}
weather_df |> 
  ggplot(aes(x = tmin, y = tmax)) + 
  geom_point(aes(color = name), alpha = .5) + 
  labs(
    title = "Temperature plot",
    x = "Minimum daily temperature (C)",
    y = "Maxiumum daily temperature (C)",
    color = "Location",
    caption = "Data from the rnoaa package") + 
  scale_color_hue(h = c(100, 300))
```
`better to use the viridis package instead for better hue colors of the graph.`

```{r}
library(viridis)
library(viridisLite)
```


```{r}
ggp_temp_plot = 
  weather_df |> 
  ggplot(aes(x = tmin, y = tmax)) + 
  geom_point(aes(color = name), alpha = .5) + 
  labs(
    title = "Temperature plot",
    x = "Minimum daily temperature (C)",
    y = "Maxiumum daily temperature (C)",
    color = "Location",
    caption = "Data from the rnoaa package"
  ) + 
  viridis::scale_color_viridis(
    name = "Location", 
    discrete = TRUE
  )
```


`Update my base plot`


```{r}
ggp_temp_plot +
  theme_minimal() +
  theme(legend.position = "bottom")
```

### Adding data in geoms

```{r}
central_park_df = 
  weather_df |>
  filter(name == "Molokai_HI")

ggplot(data = central_park_df, aes(x = date, y = tmax, color = name)) + 
  geom_point() +
  geom_line(data = central_park_df)
```


### `Patchwork`


`Make three plots and combine using patchwork.`

```{r}
ggp_tmax_tmin = 
  weather_df |>
  ggplot(aes(x = tmin, y= tmax, color = name)) +
  geom_point(alpha = 0.5)

ggp_prec_density = 
  weather_df |>
  filter(prcp > 0) |>
  ggplot(aes(x = prcp, fill = name)) +
  theme(legend.position = "bottom")

ggp_temp_season = 
  weather_df |>
  ggplot(aes(x = date, y=tmax, color = name)) +
  geom_point(alpha = 0.5)
  geom_smooth(se = FALSE)
  
  (ggp_tmax_tmin + ggp_prec_density)/ ggp_temp_season
```

### Data Manipulation


`Let's make temperature violin plots.`


```{r}
weather_df |>
  mutate(name = fct_relevel("Molokai_HI", "CentralPrak_NY", "Waterhole_WA")) |>
  ggplot(aes(x = name, y = tmax, fill = name)) + 
  geom_violin(alpha = 0.5)
```

`The above graph is supposed to have three graphs with Molokai being first one.`


```{r}
weather_df |>
  mutate(name = fct_reorder(name,tmax)) |>
  ggplot(aes(x = name, y = tmax, fill = name)) +
  geom_violin(alpha = 0.5)
```



#### ` What about data tidiness?`


```{r}
pulse_df = 
  haven::read_sas("./data/public_pulse_data.sas7bdat") |>
  janitor::clean_names() |>
  pivot_longer(
    bdi_score_bl:bdi_score_12m,
    names_to = "visit", 
    names_prefix = "bdi_score_",
    values_to = "bdi") |>
  select(id, visit, everything()) |>
  mutate(
    visit = recode(visit, "bl" = "00m"),
    visit = factor(visit, levels = str_c(c("00", "01", "06", "12"), "m"))) |>
  arrange(id, visit)

ggplot(pulse_df, aes(x = visit, y = bdi)) + 
  geom_boxplot()
```

`1. import pups dataframe

```{r}
pups_df = 
  read.csv("data/FAS_pups.csv", na = c("NA", ".",""), skip = 3) |>
  janitor::clean_names() |>
  mutate(
    sex = case_match(
      sex,
      1 ~ "male",
      2 ~ "female"
    )
  )

litters_df =
  read.csv("data/FAS_litters.csv", na = c("NA",".", "")) |>
  janitor::clean_names()|>
  separate(group, into = c ("dose", "tx_day"), sep = 3)

fas_df =
  left_join(pups_df, litters_df, by = "litter_number")

fas_df |>
  select(pd_ears:tx_day) |>
  pivot_longer(
    pd_ears:pd_walk,
    names_to = "outcome",
    names_prefix = "pd_",
    values_to = "pn_day") |>
  mutate(outcome = fct_reorder(outcome, pn_day)) |>
  drop_na() |>
  ggplot(aes(x = dose, y = pn_day)) +
  geom_violin() +
  facet_grid(tx_day ~ outcome)
```




### Learning Assessment

```{r}
weather_df |> 
  ggplot(aes(x = date, y = tmax)) + 
  geom_point(aes(color = name), alpha = .5)
  labs(
    title = "Temperature against Date",
    x = "Date",
    y = "Maxiumum daily temperature (C)",
    color = "Location",
    caption = "Data from the rnoaa package") + 
  scale_x_continuous(
    breaks = c(-20, 0, 25), 
    labels = c("-20º C", "0", "25")) +
  scale_y_continuous(
    trans = "sqrt",
    limits = c(10,30)
  )
```

